Resonant Scattering in a Uniform Sphere with Large Optical Depth

### 2021-06-21 ms.tex
adding atau1/3 scaling beginnings
committing previous changes
changing some \nu to \sigma where appropriate
Adding in new taubc plot and short description / caption

\documentclass{aastex63}

\newcommand{\be}{\begin{eqnarray}}
\newcommand{\ee}{\end{eqnarray}}
\def\la{\mathbin{\lower 3pt\hbox
      {$\rlap{\raise 5pt\hbox{$\char'074$}}\mathchar"7218$}}}
\def\ga{\mathbin{\lower 3pt\hbox
      {$\rlap{\raise 5pt\hbox{$\char'076$}}\mathchar"7218$}}} %> or of order
\renewcommand{\vec}[1]{\mathbf{#1}}
%\renewcommand{\vec}[1]{\ensuremath{\boldsymbol{#1}}} %boldface vector style
\newcommand{\grad}{\mathbf{\nabla}}
\newcommand\lya{Ly$\alpha$\ }

@@

A third class of methods are modified random walk methods \citep{1984JCoPh..54..508F,2009A&A...497..155M,Robitaille2010}. In this case a spherical volume (usually the largest within a simulation cell) is chosen when the scattering mean-free-path is small compared with the zone size.  An outgoing photon is then randomly sampled on the surface of this outgoing sphere by drawing its properties from distributions in outgoing frequencies, directions and escape times, based on solutions to the diffusion equation within a uniform sphere.

@@

A better approximation to the true solution has been derived by \citet{2006ApJ...649...14D}, who generalized the closed-form solution in slab geometry found in \citet{1990ApJ...350..216N}. They found a closed-form solution which satisfies a $J=0$ boundary condition at $r=R$. We generalize their solution to allow emission at frequency $\nu_{\rm s}$ away from line center. The result is
\be
J_0 & = & \left( \frac{\sqrt{6}L}{32\pi^2 \Delta} \right)
\left( \frac{1}{Rr} \right)
\left( 
\frac{ \sin(\pi r/R) }{ \cosh \left[ \frac{\pi \Delta}{k R} (\sigma - \sigma_{\rm s}) \right] - \cos(\pi r/R)}
\right)
\label{eq:J0}
\ee
and
\be
H_0 & = & \left( \frac{1}{3k\phi} \right)
\left( \frac{\sqrt{6}L}{32\pi^2 \Delta} \right)
\left( \frac{1}{Rr^2} \right)
\left( 
\frac{ \sin(\pi r/R) }{ \cosh \left[ \frac{\pi \Delta}{k R} (\sigma - \sigma_{\rm s}) \right] - \cos(\pi r/R)}
\right. \nonumber \\ & & \left. - \left( \frac{\pi r}{R} \right)
\frac{ \cos(\pi r/R) }{ \cosh \left[ \frac{\pi \Delta}{k R} (\sigma - \sigma_{\rm s}) \right] - \cos(\pi r/R)}
+ \left( \frac{\pi r}{R} \right)
\frac{ \sin^2(\pi r/R) }{ \left[ \cosh \left[ \frac{\pi \Delta}{k R} (\sigma - \sigma_{\rm s}) \right] - \cos(\pi r/R) \right]^2 }
\right).
\label{eq:H0}
\ee
Again $J_0 \gg H_0$, except very near the $r=R$, where it goes to zero. The flux at $r=R$ can be written
\be
H_0(R,\nu) & = & \left( \frac{1}{3k\phi} \right)
\left( \frac{\sqrt{6}L}{32\pi \Delta} \right)
\left( \frac{1}{R^3} \right)
\left( 
\frac{ 1 }{ \cosh \left[ \frac{\pi \Delta}{k R} (\sigma - \sigma_{\rm s}) \right] +1 }
\right).
\label{eq:H0surf}
\ee
Eq. \ref{eq:H0surf} will be shown to be a much better approximation to the solution, which is valid near the delta function at the center, and is a much better approximation at the surface. It will be shown that it is still only a good approximation near line center at the surface. It decreases exponentially rather than as a power-law in frequency, giving a much smaller flux on the line wings as compared to $H_{\rm d}$.

@@

A different solution method is attempted here: namely, a continuous Fourier expansion in the frequency variable $\sigma$. The full solution is written
\be
J(r,\nu) & = & J_0(r,\nu) + J_{\rm bc}(r,\nu)
\ee
and
\be
H(r,\nu) & = & H_0(r,\nu) + H_{\rm bc}(r,\nu).
\ee
Here $J_0$ and $H_0$ are the solutions of the inhomogeneous equation including the delta function source given in Equations \ref{eq:J0} and \ref{eq:H0}. The additional term $J_{\rm bc}$ must then be a solution of the homogeneous equation
\be
\frac{\partial^2J}{\partial r^2} + \frac{2}{r} \frac{\partial J}{\partial r}
+ \left( \frac{k}{\Delta} \right)^2 \frac{\partial^2 J}{\partial \sigma^2} &= & 0
\ee
with no delta function source term, and it must allow the boundary conditions to be satisfied at the surface. Since $J_0(R,
\nu)=0$, the surface boundary condition becomes
\be
J_{\rm bc}(R,\nu) - \sqrt{3} H_{\rm bc}(R,\nu) & = & 
J_{\rm bc}(R,\nu) + \frac{1}{\sqrt{3}k\phi} \frac{\partial J_{\rm bc}(R,\nu)}{\partial r} = 
\sqrt{3} H_0(R,\nu).
\label{eq:bc2}
\ee
Plugging in a frequency dependence $J \propto e^{is\sigma}$, for ``wavenumber" $s$, gives the equation for modified spherical Bessel functions of the first kind, $i_0(z)=\sinh(z)/z$.

@@

Plugging Eq. \ref{eq:Jbc} into Eq. \ref{eq:bc2} leads to the following equation for the Fourier amplitudes
\be
\int_{-\infty}^\infty \frac{ds}{2\pi} e^{is\sigma} A(s)
\left[ 1 + \left( \frac{s}{\sqrt{3} \Delta \phi} \right) \left( \frac{i_0^\prime(kRs/\Delta)}{i_0(kRs/\Delta)} \right) \right]
& = & \sqrt{3} H_0(R,\nu)
\label{eq:bc3}
\ee
where $H_0(R,\nu)$ is given by Eq. \ref{eq:H0surf}. Discretization of Eq. \ref{eq:bc3} for frequency variables $\sigma_i$ and wavenumbers $s_j$
leads to a set of coupled linear equations for the $A(s_j)$. We use equally-spaced points $\Delta \sigma = 2\sigma_{\rm max}/(N-1)$ and $\Delta s = 2\pi/(N\Delta \sigma)$, where $N$ is the number of points for each grid. The maximum frequency is set as $\sigma_{\rm max} = {\rm constant} \times \tau_0$, for a large enough constant that  the end of the frequency grid is at such small intensities that it does not affect the solution except close to the boundaries. The number of points was increased until the solution was well-resolved near line center, and only became inaccurate near the boundaries. Given the Fourier amplitudes, $J_{\rm bc}$ is computed using Eq. \ref{eq:Jbc}, and the flux is given by
\be
H_{\rm bc}(r,\nu) & = & \left( \frac{-1}{3k\phi} \right)
\frac{\partial J_{\rm bc}(r,\nu)}{\partial r}
= \left( \frac{-1}{3k\phi} \right)
\int_{-\infty}^\infty \frac{ds}{2\pi} e^{is\sigma} A(s) 
\left( \frac{ks}{\Delta} \right) 
\left( \frac{i_0^\prime(krs/\Delta)}{i_0(kRs/\Delta)} \right).
\label{eq:Hbc}
\ee
The Bessel functions are finite at the center and rise steeply toward the surface when $kRs/\Delta \gg 1$.

@@

\begin{figure}
    \centering
    \includegraphics{pdf_xinit0.0.pdf}
    \caption{Analytic solution for the outgoing spectrum compared with Monte Carlo at a line center optical depth of $\tau_0 = 1 \times 10^7$. Photons are initialized with $\rm x = 0.0$. In the log-scale plot in the second panel, $|\rm H_{bc}|$ is shown instead of $\rm H_{bc}$. Residuals of each solution to the Monte Carlo data points are shown in the third panel.} 
    \label{fig:sol_mc_residual_0}
\end{figure}

@@

 \begin{figure}
    \centering
    \includegraphics{tau_threepanel_factorof2.pdf}
    \caption{Increasing the line center optical depth. Shown here are $\tau_0 = 10^5, 10^6, 10^7$ with the $x$ and $y$ axes scaled by the characteristic frequency diffusion scale $(a\tau_0)^{1/3}$. Photons are initialized with $\rm x = 0.0$.} 
    \label{fig:sol_mc_tau}
\end{figure}

>>>>

\documentclass{aastex63}
\usepackage{amsmath}
\newcommand{\be}{\begin{eqnarray}}
\newcommand{\ee}{\end{eqnarray}}
\def\la{\mathbin{\lower 3pt\hbox
      {$\rlap{\raise 5pt\hbox{$\char'074$}}\mathchar"7218$}}}
\def\ga{\mathbin{\lower 3pt\hbox
      {$\rlap{\raise 5pt\hbox{$\char'076$}}\mathchar"7218$}}} %> or of order
\renewcommand{\vec}[1]{\mathbf{#1}}
%\renewcommand{\vec}[1]{\ensuremath{\boldsymbol{#1}}} %boldface vector style
\newcommand{\grad}{\mathbf{\nabla}}
\newcommand\lya{Ly$\alpha$\ }

@@

A third class of methods are modified random walk methods \citep{1984JCoPh..54..508F,2009A&A...497..155M,Robitaille2010}. In this case a spherical volume (usually the largest within a simulation cell) is chosen when the scattering mean-free-path is small compared with the zone size.  An outgoing photon is then randomly sampled on the surface of this outgoing sphere by drawing its properties from distributions in outgoing frequencies, directions and escape times, based on solutions to the diffusion equation within a uniform sphere. A method similar to this has been applied by \citet{2006ApJ...645..792T} to Lyman $\alpha$ transfer using the \cite{1990ApJ...350..216N} solution.

@@

A better approximation to the true solution has been derived by \citet{2006ApJ...649...14D}, who generalized the closed-form solution in slab geometry found in \citet{1990ApJ...350..216N}. They found a closed-form solution which satisfies a $J=0$ boundary condition at $r=R$. We generalize their solution to allow emission at frequency $\nu_{\rm s}$ away from line center. The result is
\be
J_0 & = & \left( \frac{\sqrt{6}L}{32\pi^2 \Delta} \right)
\left( \frac{1}{Rr} \right)
\left( 
\frac{ \sin(\pi r/R) }{ \cosh \left[ \frac{\pi \Delta}{k R} (\sigma - \sigma_{\rm s}) \right] - \cos(\pi r/R)}
\right)
\label{eq:J0}
\ee
and
\be
H_0 & = & \left( \frac{1}{3k\phi} \right)
\left( \frac{\sqrt{6}L}{32\pi^2 \Delta} \right)
\left( \frac{1}{Rr^2} \right)
\left( 
\frac{ \sin(\pi r/R) }{ \cosh \left[ \frac{\pi \Delta}{k R} (\sigma - \sigma_{\rm s}) \right] - \cos(\pi r/R)}
\right. \nonumber \\ & & \left. - \left( \frac{\pi r}{R} \right)
\frac{ \cos(\pi r/R) }{ \cosh \left[ \frac{\pi \Delta}{k R} (\sigma - \sigma_{\rm s}) \right] - \cos(\pi r/R)}
+ \left( \frac{\pi r}{R} \right)
\frac{ \sin^2(\pi r/R) }{ \left[ \cosh \left[ \frac{\pi \Delta}{k R} (\sigma - \sigma_{\rm s}) \right] - \cos(\pi r/R) \right]^2 }
\right).
\label{eq:H0}
\ee
Again $J_0 \gg H_0$, except very near the $r=R$, where it goes to zero. The flux at $r=R$ can be written
\be
H_0(R, \sigma) & = & \left( \frac{1}{3k\phi} \right)
\left( \frac{\sqrt{6}L}{32\pi \Delta} \right)
\left( \frac{1}{R^3} \right)
\left( 
\frac{ 1 }{ \cosh \left[ \frac{\pi \Delta}{k R} (\sigma - \sigma_{\rm s}) \right] +1 }
\right).
\label{eq:H0surf}
\ee
Eq. \ref{eq:H0surf} will be shown to be a much better approximation to the solution, which is valid near the delta function at the center, and is a much better approximation at the surface. It will be shown that it is still only a good approximation near line center at the surface. It decreases exponentially rather than as a power-law in frequency, giving a much smaller flux on the line wings as compared to $H_{\rm d}$.

@@

A different solution method is attempted here: namely, a continuous Fourier expansion in the frequency variable $\sigma$. The full solution is written
\be
J(r,\nu) & = & J_0(r,\nu) + J_{\rm bc}(r,\nu)
\ee
and
\be \label{eq:totalflux}
H(r,\nu) & = & H_0(r,\nu) + H_{\rm bc}(r,\nu).
\ee
Here $J_0$ and $H_0$ are the solutions of the inhomogeneous equation including the delta function source given in Equations \ref{eq:J0} and \ref{eq:H0}. The additional term $J_{\rm bc}$ must then be a solution of the homogeneous equation
\be
\frac{\partial^2J}{\partial r^2} + \frac{2}{r} \frac{\partial J}{\partial r}
+ \left( \frac{k}{\Delta} \right)^2 \frac{\partial^2 J}{\partial \sigma^2} &= & 0
\ee
with no delta function source term, and it must allow the boundary conditions to be satisfied at the surface. Since $J_0(R,
\nu)=0$, the surface boundary condition becomes
\be
J_{\rm bc}(R,\nu) - \sqrt{3} H_{\rm bc}(R,\nu) & = & 
J_{0}(R,\nu) + \frac{1}{\sqrt{3}k\phi} \frac{\partial J_{0}(R,\nu)}{\partial r} = 
\sqrt{3} H_0(R,\nu).
\label{eq:bc2}
\ee
Plugging in a frequency dependence $J \propto e^{is\sigma}$, for ``wavenumber" $s$, gives the equation for modified spherical Bessel functions of the first kind, $i_0(z)=\sinh(z)/z$.

@@

Plugging Eq. \ref{eq:Jbc} into Eq. \ref{eq:bc2} leads to the following equation for the Fourier amplitudes
\be
\int_{-\infty}^\infty \frac{ds}{2\pi} e^{is\sigma} A(s)
\left[ 1 + \left( \frac{s}{\sqrt{3} \Delta \phi} \right) \left( \frac{i_0^\prime(kRs/\Delta)}{i_0(kRs/\Delta)} \right) \right]
& = & \sqrt{3} H_0(R,\nu)
\label{eq:bc3}
\ee
where $H_0(R,\nu)$ is given by Eq. \ref{eq:H0surf}. Discretization of Eq. \ref{eq:bc3} for frequency variables $\sigma_i$ and wavenumbers $s_j$
leads to a set of coupled linear equations for the $A(s_j)$. We use equally-spaced points $\Delta \sigma = 2\sigma_{\rm max}/(N-1)$ and $\Delta s = 2\pi/(N\Delta \sigma)$, where $N$ is the number of points for each grid. The maximum frequency is set as $\sigma_{\rm max} = {\rm constant} \times \tau_0$, for a large enough constant that  the end of the frequency grid is at such small intensities that it does not affect the solution except close to the boundaries. \todo{The number of points was increased until the solution was well-resolved near line center, and only became inaccurate near the boundaries. (Is this still accurate?)} Given the Fourier amplitudes, $J_{\rm bc}$ is computed using Eq. \ref{eq:Jbc}, and the flux is given by
\be
H_{\rm bc}(r,\nu) & = & \left( \frac{-1}{3k\phi} \right)
\frac{\partial J_{\rm bc}(r,\nu)}{\partial r}
= \left( \frac{-1}{3k\phi} \right)
\int_{-\infty}^\infty \frac{ds}{2\pi} e^{is\sigma} A(s) 
\left( \frac{ks}{\Delta} \right) 
\left( \frac{i_0^\prime(krs/\Delta)}{i_0(kRs/\Delta)} \right).
\label{eq:Hbc}
\ee
The Bessel functions are finite at the center and rise steeply toward the surface when $kRs/\Delta \gg 1$.

To examine the relative size of this corrective term as compared to $H_0$, we consider the case where the optical depth is large and thus the mean intensity will be considerably larger than the surface flux. Starting from Eq. \ref{eq:bc2}, we assume $H_{\rm bc} \ll J_{\rm bc}$ such that
\be
J_{\rm bc}(R,\sigma) & \approx & \sqrt{3} H_0(R,\sigma).
\label{eq:bc4}
\ee
The Fourier transform for $J_{\rm bc}$ at the surface becomes
\be
J_{\rm bc}(R, \sigma) \equiv \int_{-\infty}^{\infty} \frac{ds}{2\pi} e^{is\sigma} A(s)
\ee
with Fourier amplitudes
\be \label{eq:amps}
A(s) = \sqrt{3} \int_{-\infty}^{\infty} d\sigma e^{-is\sigma} H_0(R, \sigma).
\ee
Plugging the expression for $H_0$ in Eq. \ref{eq:H0surf} into Eq. \ref{eq:amps} yields an expression for the Fourier amplitudes
\be \label{eq:plugged_amps}
A(s) = \sqrt{3}\frac{1}{3k}\frac{\sqrt{6}}{32\pi}\frac{L}{\Delta R^3} \int_{-\infty}^\infty d\sigma e^{-is\sigma}\frac{1}{\phi(\sigma)}\frac{1}{1+\cosh \left[\frac{\pi \Delta}{kR} |\sigma - \sigma_i|\right]}
\ee
with
\be
\phi(\sigma) = \left(\frac{2a}{27\pi}\right)^{1/3}\frac{1}{\Delta |\sigma|^{2/3}}.
\ee
With the line center optical depth defined as
\be
\tau_0 = \frac{kR}{\sqrt{\pi}\Delta}
\ee
we scale the variables $\sigma$ and $s$ by $\tau_0$, now using characteristic photon frequency $\bar \sigma = \sigma / \tau_0$ and characteristic wavenumber $\bar s = s \tau_0$. Eq. \ref{eq:plugged_amps} becomes
\be
A(\bar s) &=& \frac{\sqrt{2}}{32\pi}\frac{L}{kR^3}\int_{-\infty}^\infty d\bar\sigma \tau_0 e^{-i\bar s \bar\sigma} \left(\frac{27\pi}{2a}\right)^{1/3}\tau_0^{2/3}|\bar\sigma^{2/3}|\frac{1}{1+\cosh \left[\sqrt{\pi}|\bar\sigma - \bar\sigma_i|\right]}\\
&=& \frac{\sqrt{2}}{32\pi}\left(\frac{27\pi}{2}\right)^{1/3}\frac{L}{kR^3}\frac{\tau_0^{5/3}}{a^{1/3}}\int_{-\infty}^\infty d\bar\sigma e^{-i\bar s \bar\sigma} |\bar\sigma^{2/3}|\frac{1}{1+\cosh \left[\sqrt{\pi}|\bar\sigma - \bar\sigma_i|\right]}
\ee
These coefficients are used to evaluate Eq. \ref{eq:Hbc} at the surface of the sphere. Dividing this flux $H_{\rm bc}$ by the $H_0$ (Eq. \ref{eq:H0surf}) yields an overall proportionality
\be
\frac{H_{\rm bc}(R, \sigma)}{H_0(R, \sigma)} \propto \frac{1}{(a\tau_0)^{1/3}}.
\ee
Thus, the fractional importance of the corrective frequency-dependent boundary condition term scales as the inverse cube root of the damping parameter $a$ times the line-center optical depth $\tau_0$. At very large optical depths, it is expected that the corrective term will be small, but will become increasingly important in less optically thick media. This is shown graphically in Fig. \ref{fig:taubc}, where the $H_{\rm bc}$ solution has been discretized in $s$ and $\sigma$ and solved via a matrix equation.

\begin{figure}
    \centering
    \includegraphics{taubc.pdf}
    \caption{Boundary condition term $H_{\rm bc}$ scaled by $(a\tau_0)^{1/3}$ shown for increasing line center optical depths. The $H_0$ solution minus the Monte Carlo data should be close to $H_{\rm bc}$ according to Eq. \ref{eq:totalflux}. As $\tau_0$ becomes larger, the agreement between these two improves, indicating that at large optical depths the relative size of the corrective boundary condition term scales as $(a\tau_0)^{-1/3}$}.
    \label{fig:taubc}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics{final_residual.pdf}
    \caption{Analytic solution for the outgoing spectrum compared with Monte Carlo at a line center optical depth of $\tau_0 = 1 \times 10^7$. Photons are initialized with $\rm x = 0.0$. In the log-scale plot in the second panel, $|\rm H_{bc}|$ is shown instead of $\rm H_{bc}$. Residuals of each solution to the Monte Carlo data points are shown in the third panel.} 
    \label{fig:sol_mc_residual_0}
\end{figure}

@@

 \begin{figure}
    \centering
    \includegraphics{tau_threepanel.pdf}
    \caption{Increasing the line center optical depth. Shown here are $\tau_0 = 10^5, 10^6, 10^7$ with the $x$ and $y$ axes scaled by the characteristic frequency diffusion scale $(a\tau_0)^{1/3}$. Photons are initialized with $\rm x = 0.0$.} 
    \label{fig:sol_mc_tau}
\end{figure}
